---
title: "Example code for ggplot2"
author: "Kevin Rue-Albrecht/modified by Carla Cohen"
date: "12/05/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(ggrepel)
library(cowplot)
library(patchwork)
library(gridExtra)
```

# Exercise

Use `library(ggplot2)` to load the package.

```{r}
library(ggplot2)
```

## Specify the dataset.

What does `ggplot(diamonds)` do?

```{r}
ggplot(diamonds)
```

> Answer:

Nothing because you haven't said what you want to plot

## Add the aesthetics.

What does `ggplot(diamonds, aes(x = carat, y = price))` do?

```{r}
ggplot(diamonds, aes(x = carat, y = price))
```

> Answer:
We have defined the range of the axes
But we still haven't plotted anything

## Add geometric objects

- Add data points showing `carat` on the x-axis and `price` on the y-axis.

```{r}
ggplot(diamonds, aes(x = carat, y = price))+
  geom_point()
  
```

- Color data points by `cut`.
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(diamonds, aes(x = carat, y = price, colour = cut))+
  geom_point()
  
```
NB the legend was added automatically because we added something that was not on the x or y axis.

- Add a smoothed mean trend line.
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(diamonds, aes(x = carat, y = price, colour = cut))+
  geom_point()+
  geom_smooth()


```

- Assign that last plot to an object called `obds_diamonds`.

```{r}

obds_diamonds <- ggplot(diamonds, aes(x = carat, y = price, colour = cut))+
  geom_point()+
  geom_smooth()
obds_diamonds

```


# Exercise

Predict the difference between these two plots

### Plot 1

```{r, include=TRUE}
ggplot(diamonds, aes(x = carat, y = price, colour = cut)) +
  geom_point() +
  geom_smooth()
```

### Plot 2

```{r, include=TRUE}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(aes(colour = cut)) +
  geom_smooth()
```

> Answer:
When you put colour = cut into the ggplot command, it is applicable to all the subsequent layers. 
But when you put colour = cut into the geom_point command, it is only applicable to that one and the geom_smooth then considers the whole data set, not each colour separately. 

# Exercise

## Trend lines

Using the `ChickWeight` data set:

- Create a scatter plot of weight (y-axis) over time (x-axis).

```{r}
data("ChickWeight")
head(ChickWeight)

ggplot(ChickWeight, aes(x = Time, y = weight))+
  geom_point()

```

- Color by diet.
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(ChickWeight, aes(x = Time, y = weight, colour = Diet))+
  geom_point()

```

- Add a linear mean trend line for each diet.
  (Copy-paste and extend the code chunk above.)

```{r}

ggplot(ChickWeight, aes(x = Time, y = weight, colour = Diet))+
  geom_point()+
  geom_smooth(method = "lm")

```

At this point you should be able to visually identify
the diet that leads to the highest mean increase in weight.

> Answer:
Diet 3

- Facet a ribbon of sub-plots, one per diet.
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(ChickWeight, aes(x = Time, y = weight, colour = Diet))+
  geom_point()+
  geom_smooth(method = "lm")+
  facet_wrap(~Diet)



```
This makes it more obvious to see that there is an outlier in Diet 2 where one chick has not increased in weight.

- Assign that last plot to an object called `obds_chickweight`.

```{r}
obds_chickweight <- ggplot(ChickWeight, aes(x = Time, y = weight, colour = Diet))+
  geom_point()+
  geom_smooth(method = "lm")+
  facet_wrap(~Diet)



```

# Exercise

## Bar plot

- Load the `ggplot2::msleep` data set.

```{r}
head(msleep)

```

- Draw a bar plot of number of observations (i.e., rows) for each taxonomic order
  (i.e, one plot and one bar per taxonomic order).

```{r}
ggplot(msleep, aes(x = order))+
  geom_bar()

```

Need to use geom_bar() instead of gemo_histogram() because it is a discreet not continuous variable.

- Change the angle and font size of the text for the x-axis ticks (not the axis titles).
  Justify the text of those x-axis ticks as right-aligned.
  (Copy-paste and extend the code chunk above.)

```{r}

ggplot(msleep, aes(x = order))+
  geom_bar()+
  theme(axis.text.x = element_text(angle = 90, size = 12, hjust = 1, vjust = 0.5))

```

- Change the value and font size of the title for both x and y axes.
  (Copy-paste and extend the code chunk above.)

```{r}

ggplot(msleep, aes(x = order))+
  geom_bar()+
  theme(axis.text.x = element_text(angle = 90, size = 12, hjust = 1, vjust = 0.5))+
  labs(x = "Taxonomic Order", y = "Number of Animals", title = "msleep dataset")+
  theme(axis.title = element_text(size = 14))

```

- Fill each bar with colors, proportionally to the count of each genus.
  (Copy-paste and extend the code chunk above.)

From this point onwards, you may need to iteratively resize the text of the ticks and axes for readability.

```{r}
head(msleep$genus)

ggplot(msleep, aes(x = order, fill = genus))+ #colour = genus would alter the line around the bar
  geom_bar()+
  theme(axis.text.x = element_text(angle = 90, size = 12, hjust = 1, vjust = 0.5))+
  labs(x = "Taxonomic Order", y = "Number of Animals", title = "msleep dataset")+
  theme(axis.title = element_text(size = 14))


```

- Reduce the legend key size.
  (Recommendation: use `unit(2, "mm")`).
  (Copy-paste and extend the code chunk above.)

```{r}

ggplot(msleep, aes(x = order, fill = genus))+ #colour = genus would alter the line around the bar
  geom_bar()+
  theme(axis.text.x = element_text(angle = 90, size = 12, hjust = 1, vjust = 0.5), 
        legend.key.size = unit(2, "mm"), axis.title = element_text(size = 14))+
  labs(x = "Taxonomic Order", y = "Number of Animals", title = "msleep dataset")

```

- Force the legend to be display in 3 columns.
  (Recommendation: use `guide_legend(...)`).
  (Copy-paste and extend the code chunk above.)

```{r}


ggplot(msleep, aes(x = order, fill = genus))+ #colour = genus would alter the line around the bar
  geom_bar()+
  theme(axis.text.x = element_text(angle = 90, size = 12, hjust = 1, vjust = 0.5), 
        legend.key.size = unit(2, "mm"), axis.title = element_text(size = 14))+
  labs(x = "Taxonomic Order", y = "Number of Animals", title = "msleep dataset")+
  guides(fill = guide_legend(ncol = 3)) #we use fill here because this is related to the fill legend (genus)

```

- Add a contour of thin black lines to the bars.

```{r}


ggplot(msleep, aes(x = order, fill = genus))+ #colour = genus would alter the line around the bar
  geom_bar(colour = "black", size = 0.1)+ #colour = black goes here because it is not related to the data so not an aesthetic
  theme(axis.text.x = element_text(angle = 90, size = 12, hjust = 1, vjust = 0.5), 
        legend.key.size = unit(2, "mm"), axis.title = element_text(size = 14))+
  labs(x = "Taxonomic Order", y = "Number of Animals", title = "msleep dataset")+
  guides(fill = guide_legend(ncol = 3)) #we use fill here because this is related to the fill legend (genus)

```

- Assign that last plot to an object called `obds_msleep`.

```{r}

obds_msleep <- ggplot(msleep, aes(x = order, fill = genus))+ #colour = genus would alter the line around the bar
  geom_bar(colour = "black", size = 0.1)+ #colour = black goes here because it is not related to the data so not an aesthetic
  theme(axis.text.x = element_text(angle = 90, size = 12, hjust = 1, vjust = 0.5), 
        legend.key.size = unit(2, "mm"), axis.title = element_text(size = 14))+
  labs(x = "Taxonomic Order", y = "Number of Animals", title = "msleep dataset")+
  guides(fill = guide_legend(ncol = 3)) #we use fill here because this is related to the fill legend (genus)

```

# Exercise

## Plotting grid

- Collate the plots that we assigend to objects through the day, as a single plot.

  + Plots: `obds_diamonds`, `obds_chickweight`, `obds_msleep`.
  
  + Methods: `cowplot::plot_grid()`, `patchwork`, `gridExtra::grid.arrange()`.

Using `cowplot`.

Aim to put diamonds and chickweight on top row and msleep on the bottom row

```{r}


library(cowplot)
#with cowplot we need to do it in two steps in order to generate the correct orientation.

first_row <- plot_grid(obds_chickweight, obds_diamonds, labels = c("A","B"), ncol = 2, nrow =1)
second_row <- plot_grid(obds_msleep, labels = c("C"), ncol = 1, nrow =1)
super_plot <- plot_grid(first_row, second_row, ncol = 1, nrow = 2); super_plot
```

Using `patchwork`.

```{r, fig.height=12, fig.width=12}
library(patchwork)
my_plots <- (obds_chickweight + obds_diamonds) / obds_msleep
my_plots

```

Using `gridExtra`.
See this page for more details
https://cran.r-project.org/web/packages/gridExtra/vignettes/arrangeGrob.html
A grob means a plot.
In one line we need to give the entire layout.
Enables complex layouts, using a matrix that defines the layout. 
```{r}
#create the matrix

lay <- rbind(c(1,2), c(3,3))
```
Then do the layout.
This creates an object that is using baseR not ggplot so the save is different.
Also you cannot save it to a variable.

```{r, fig.height=12, fig.width=12}
library(gridExtra)
grid.arrange(obds_diamonds,obds_chickweight, obds_msleep, layout_matrix=lay)

```

- Export the new plot in a PDF file, and open it in a PDF viewer (e.g. Adobe Acrobat Reader DC).

You will likely need a few attempts to fine-tune the width and height of the output file.

```{r}
ggsave("patchwork_plot.pdf", plot = my_plots, width = 15, height = 10)
ggsave("cowplot_plot.pdf", plot = super_plot, width = 15, height = 10)

```
There are minor differences in the layout, e.g. in terms of how the plots are aligned to one another. 


Now use a different method to save the gridextra_plot because it is not a ggplot output so we cannot use ggsave.
```{r}
pdf(file = "gridextra_plot.pdf", width = 15, height = 10) #open the pdf
#write the plot to the device
grid.arrange(obds_diamonds,obds_chickweight, obds_msleep, layout_matrix=lay)
dev.off()
```

# Exercise

## Pair programming

- Explore the data set `ggplot2::mpg` and generate the most informative plot that you can!

Added a line y = x i.e. if the mpg was the same in the city as in the highway.
```{r}
ggplot(mpg, aes(x = cty, y = hwy, colour = year))+
  geom_point(size = 0.5)+ #change the size of the point
  facet_wrap(~manufacturer)
```


```{r}
head(mpg)

ggplot(mpg, aes(x = cty, y = hwy, colour = year))+
  geom_point(size = 0.5)+ #change the size of the point
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", size = 0.5)+ #add a line of y = x
  facet_wrap(~manufacturer)+
  theme(panel.grid.minor = element_blank(),#to delete an element use element_blank(), this gets rid of the major and minor grid lines
        panel.grid.major = element_blank(),
        strip.text = element_text(size = 10),#change the size of the car manufacturer facet titles
        plot.title = element_text(size = 16, hjust = 0.5))+ #change overall plot title text size
  labs(x = "City Petrol Usage", #labels go in the labs command
       y = "Highway Petrol Usage",
       title = "Main title",
       subtitle = "This is the plot subtitle")
  
```
Another example is to plot multiple layers from two different df into the same plot.

First create some fake data

```{r}
df1 <- data.frame(col1 = c(1, 2, 3), col2 = c(6, 5, 4))
df2 <- data.frame(column1 = c(1.5, 1.75), column2 = c(3.5, 4))
df3 <- data.frame(variable1 = runif(100, min = 1, max = 3))#generate some random numbers
head(df3)

```

Then make a plot.
The order that they are written will be the order of the layers. 
Make them transparent using alpha = 0.5
For histogram specify bin width with binwidth or number of bins with bins.
For geom_point, the shape can be set using a number (square, triangle etc, see the legend on the cheatsheet).
You can add labels using geom_label.


```{r}
ggplot() +
  geom_histogram(aes(x = variable1), df3, alpha = 0.5, fill = "green") + #specify the df as well as the variable
  geom_point(aes(x = col1, y = col2), df1, color = "blue") +
  geom_point(aes(x = column1, y = column2), df2, color = "red")+
  geom_label(aes(x = column1, y = column2, label = column1), df2, colour = "red")
```


