---
title: "Tidyverse exercises"
author: "Kevin Rue-Albrecht/modified by Carla Cohen"
date: "13/05/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Demo

- Load the tidyverse

```{r}
library(tidyverse)
```

- Load a single package from the tidyverse

```{r}
library(ggplot2)
```

# Demo

# The pipe operator

```{r}
x <- 4
x %>% sqrt()
```

```{r}
x <- 4
sqrt(x)
```

# Demo

## The tidyverse philosophy

```{r}
iris %>%
    select(Sepal.Length, Sepal.Width, Species) %>%
    slice_head(n = 3)
```

# Exercise

## Read and write files

- Read data from the file `iris.csv`.
  Assign the data imported from the file to an object called `iris_raw`.

```{r}
iris_raw <- read_csv("/project/obds/shared/resources/2_r/r_tidyverse/iris.csv")
```

**What do you learn about the data from the messages displayed in the R console while the contents of the file are parsed and imported into your R session?**

> Answer:
It is a tibble with 150 rows and 5 columns
<dbl> means double (number with a decimal place)
<chr> means character
A tibble tells you the type of each column (which a df does not)
Use spec(tibble) to show the list of column types.
Also a tibble has the description in the first row.
It automatically only prints first 10 rows, you don't have to use head.
Delimiter is "," (as we used read_csv)



- Print the value of `iris_raw`.

```{r}
print(iris_raw)
```

**What is the class of the data? What are the dimensions of the dataset? What is the type of data stored in each column?**

> Answer:
See above

- Write the dataset to a file named `iris.tsv`, separating fields with the tabulation character.

```{r}
write_tsv(iris_raw, "iris.tsv")

```

**What function do you use? What options are available for that function?**

> Answer:

- Inspect the `iris.tsv` file.
  You can use `file.edit()` to open the file in the RStudio editor.

```{r, eval=FALSE}
file.edit("iris.tsv")
```

**Are you satisfied with the contents and appearance of the file?**

> Answer:

# Demo

## Making a tibble

```{r}
tibble(x = 1:5, y = 1, z = x ^ 2 + y)
```

# Demo

## Subset the columns of a table

```{r}
iris %>%
    select(Sepal.Length, Sepal.Width)
```

```{r}
iris %>%
    select(starts_with("Petal") | ends_with("Width"))
```

```{r}
iris %>%
    select(!ends_with("Width"))
```

```{r}
iris %>%
    select(!c(Sepal.Length, Petal.Length))
```

# Demo

## Create and update columns in a table

```{r}
iris %>%
    mutate(
        ID = seq(1, nrow(iris)),
        Flower.ID = paste0(Species, ID)
        ) %>%
    slice_head()
```

# Demo

## Subset observations in a table

```{r}
iris %>%
    filter(Sepal.Length > mean(Sepal.Length) & Sepal.Width > mean(Sepal.Width)) %>%
    as_tibble()
```

# Demo

## Compute summary statistics

Without grouping

```{r}
iris %>%
    summarise(Sepal.Length.mean = mean(Sepal.Length))
```

With grouping

```{r}
iris %>%
    group_by(Species) %>%
    summarise(Sepal.Length.mean = mean(Sepal.Length))
```

# Demo

## Sort observations

```{r}
iris %>%
    arrange(Species, desc(Sepal.Length)) %>%
    as_tibble()
```

# Demo

## Extract a single column as a vector

Without names

```{r}
iris %>%
    pull(Sepal.Length) %>%
    head(5)
```

With names

```{r}
iris %>%
    pull(Sepal.Length, name = Species) %>%
    head(5)
```

# Demo

## Combine two tables using shared information

```{r}
tibble_1 <- tibble(
  ID = paste0("sample", 1:4),
  gene1 = rbinom(4, 10, 0.5),
  gene2 = rbinom(4, 10, 0.5)
)
tibble_1
```

```{r}
tibble_2 <- tibble(
  ID = paste0("sample", 1:4),
  batch = factor(rep(c("A", "B"), each = 2)),
  condition = factor(rep(c("control", "treated"), times = 2)),
)
tibble_2
```

**How would you describe how to join these two tibbles?**

```{r}
tibble_joined <- left_join(tibble_1, tibble_2, by = "ID")
tibble_joined
```

# Exercise

## Manipulate data

### Exercise 1

- Using `iris_raw`, for each species of iris, compute the following summary statistics for the `sepal_length`: mean, median, minimum, maximum.

summarise(Sepal.Length.mean = mean(Sepal.Length))

First use group_by to get the species, then use summarise to apply the summary stats. 

```{r}
iris_raw %>% group_by(species) %>%
  summarise(mean = mean(sepal_length), 
  median = median(sepal_length), 
  minimum = min(sepal_length),
  maximum = max(sepal_length))

```

### Exercise 2

- For each species of iris, compute the mean of every column that is numeric.
  **Hint:** use the functions `dplyr::across()`, `tidyselect::where()`, and `base::is.numeric()`.

```{r}
iris_raw %>% group_by(species) %>%
  summarise(across(where(is.numeric), mean))
    
```

Try another example: summarise columns that start with "sepal"
We are using the across function to select which columns to summarise. 
Putting things on new lines helps you work out what is the argument for what

```{r}
iris_raw %>% group_by(species) %>%
  summarise(
    across(
      .cols = starts_with("sepal"), 
      .fns = mean
      )
    )

```



- Filter the table above to retain only species of iris with an average sepal length less than `6`.

```{r}
# Copy the code chunk above and extend with more pipes
iris_raw %>% group_by(species) %>%
  summarise(across(where(is.numeric), mean)) %>%
  filter(sepal_length < 6)
    
```

- Sort the table above by descending `sepal_length`.

```{r}
# Copy the code chunk above and extend with more pipes
iris_raw %>% group_by(species) %>%
  summarise(across(where(is.numeric), mean)) %>%
  filter(sepal_length < 6) %>%
  arrange(desc(sepal_length))
    
```

- From the table above, extract the `sepal_length` column as a numeric vector.
  Make it a named numeric vector, where each value is named with the corresponding species.

```{r}
# Copy the code chunk above and extend with more pipes
iris_raw %>% group_by(species) %>%
  summarise(across(where(is.numeric), mean)) %>%
  filter(sepal_length < 6) %>%
  arrange(desc(sepal_length)) %>%
  pull(sepal_length, name = species)
    
```

# Exercise

## Manipulate data

### Exercise 3

Let's make the silly assumption that iris sepals are rectangular in shape.

- Using `iris_raw`, compute a new column named `sepal_area`, which is the product of `sepal_length` and `sepal_width`.

```{r}
iris_raw %>% mutate(sepal_area = sepal_length * sepal_width)
    
```

- Subset the result to the columns named `species` and `sepal_area`.

```{r}
# Copy the code chunk above and extend with more pipes
iris_raw %>% mutate(sepal_area = sepal_length * sepal_width) %>%
  select(species, sepal_area)
    
```

- Subset the result to display the top 5 observations by `sepal_area` (by largest value).

```{r}
# Copy the code chunk above and extend with more pipes
iris_raw %>% mutate(sepal_area = sepal_length * sepal_width) %>%
  select(species, sepal_area)%>%
  slice_max(sepal_area, n=5)
    
```

## Bonus point

- Make a histogram of `sepal_area` colored by species.

You might also want to facet the plot by species.

```{r}
# Copy the code chunk above and extend with more pipes
iris_for_ggplot <- iris_raw %>% mutate(sepal_area = sepal_length * sepal_width) %>%
  select(species, sepal_area)
    
iris_for_ggplot
```

```{r}
ggplot(iris_for_ggplot, aes(sepal_area, fill = species)) +
  geom_histogram()+
  facet_wrap(~species, ncol = 2)
    
```

# Exercise

## Pivot data from wide to long

Reshape the `iris_raw` dataset in a tidy format where one observation is represented by:

- the species

- the variable measured

- the value

**Hint:** you want to pivot all the columns that start are numeric.

```{r}
iris_raw
iris_long <- iris_raw %>% pivot_longer(1:4, names_to = "variable")
#alternative to select numeric columns specifically
iris_raw %>% pivot_longer(cols = where(is.numeric), names_to = "variable")
    
iris_long
```

**What information have we lost in the process? What could we do to remedy the issue?**

> Answer:
We have lost the detail of which information has come from which flower, so you could not pivot this table back to a wide format.

So we need to add a column that contains a flower identifier. 
```{r}
# Copy the code chunk above and refine to address the issue
iris_long2 <- iris_raw %>% 
  mutate(flower_ID = as.character(seq(1:nrow(iris_raw)))) %>% #we add as.character so that is not a numeric
  pivot_longer(cols = where(is.numeric), names_to = "variable")
    
iris_long2
```

# Exercise

## Pivot data from long to wide

- Reshape the tidy format of the iris data set into the original wide format.

**Hint:** you will only be able to restore the wide format if you kept track of the identity of each flower in the long format.

```{r}
# Copy the code chunk above and refine to address the issue
iris_wide <- iris_long2 %>% 
  pivot_wider(names_from = variable, values_from = value) %>%
  select(-flower_ID)
  
iris_wide
```

# Demo

## Split a column value into multiple columns

```{r}
iris %>% 
    separate(Sepal.Length, c("Sepal.Length.unit", "Sepal.Length.decimal"), sep = "[.]") %>%
    select(c("Sepal.Length.unit", "Sepal.Length.decimal")) %>%
    as_tibble()
```

# Demo

## Combine multiple columns into a single value

```{r}
iris %>% 
  mutate(ID = seq(1, nrow(iris))) %>% 
  unite("FlowerID", Species, ID, sep = "_") %>% 
  as_tibble()
```

# Demo

## Extract substrings
pull makes a vector of the species column

```{r}
iris_species <- iris %>%
    pull(Species)
iris_species
```

```{r}
iris_species %>%
    str_sub(1, 3) %>% #pull out positions 1-3
    unique()
```

```{r}
str_sub(iris_species, 4) <- "..."
iris_species %>%
    unique()
```

# Demo

## Join multiple strings and remove whitespaces

```{r}
words <- c("A ", " few ", "words")
words
```

```{r}
words %>%
    str_trim()
```

```{r}
words %>%
    str_trim() %>%
    str_c(collapse = " ")
```

# Integrated exercise

## Gene expression

### Part 1

We provide you with an Excel file that contains gene expression data for the nine most highly variable genes in a pancreas cancer gene expression dataset (<https://portal.gdc.cancer.gov/projects/TCGA-PAAD>), as well as metadata for the individual samples and genes.

- Import the gene expression matrix from the Excel file into the R session as a `tibble`.
  Assign it to an object named `exprs_assay`.

```{r}
library(readxl) #not loaded automatically as not a core tidyverse package
excel_sheets("/project/obds/shared/resources/2_r/r_tidyverse/EH2011.xlsx") #list all sheets in an excel spreadsheet

#read in the excel spreadsheet and specify the sheet
exprs_assay <- read_excel("/project/obds/shared/resources/2_r/r_tidyverse/EH2011.xlsx", sheet = "exprs")
    
exprs_assay
```

- Reshape the gene expression matrix in a tidy format.
  Assign it to a new object `exprs_assay_long`.

```{r}
exprs_assay_long <- exprs_assay %>% 
  pivot_longer(cols = !gene, names_to = "ProjectID")
#could also do cols = starts_with("TCGA")
    
exprs_assay_long
```

- Draw a histogram of gene expression levels for each gene in a separate facet.

```{r}
ggplot(exprs_assay_long, aes(value, fill = gene)) +
  geom_histogram()+
  facet_wrap(~gene, ncol = 3)+
  theme_classic()+
  theme(strip.background = element_blank(), strip.text = element_text(face = "italic"))

```

---

# Integrated exercise

## Gene expression

### Part 2

- Import the sample metadata from the Excel file into the R session as a `tibble`.
  Assign it to an object named `sample_info`.

```{r}
excel_sheets("/project/obds/shared/resources/2_r/r_tidyverse/EH2011.xlsx") #list all sheets in an excel spreadsheet
sample_info <- read_excel("/project/obds/shared/resources/2_r/r_tidyverse/EH2011.xlsx", sheet = "sample_info")
sample_info
```

> Metadata key:
> 
> [NIH AJCC staging system](https://www.cancer.gov/publications/dictionaries/ca> ncer-terms/def/ajcc-staging-system)
> 
> - grade: how abnormal the cancer cells and tissue look under a microscope
> - T: size of the tumor and any spread of cancer into nearby tissue
> - N: spread of cancer to nearby lymph nodes
> - M: metastasis (spread of cancer to other parts of the body)

- Convert the four columns above to the factor type.

```{r}
sample_info <- sample_info %>%
  mutate(grade=as.factor(grade), T=as.factor(T), N=as.factor(N), M=as.factor(M))

sample_info_2 <- sample_info %>%
  mutate(
    across(
    .cols = c(grade, T, N, M),
    .fns = as.factor
    )
  )
    
sample_info
sample_info_2
```

- Combine sample information with gene expression data.
  Assign the output to `exprs_sampleinfo`.

```{r}
exprs_assay_long

exprs_sampleinfo <- sample_info %>%
  left_join (exprs_assay_long, by = c("unique_patient_ID" = "ProjectID"))

#could also have renamed the column in the table rather than using the by = argument.
    
exprs_sampleinfo
```

---

# Integrated exercise

## Gene expression

### Part 3

- Identify the gene with the largest absolute difference in expression between male and female patients.

```{r}
exprs_sampleinfo

exprs_sampleinfo %>% 
  group_by(gene, sex) %>%
  summarise(gene_mean = mean(value)) %>%
  pivot_wider(names_from = sex, values_from = gene_mean) %>%
  mutate (difference = F - M) %>%
  arrange(desc(abs(difference))) #abs means don't worry if positive or negative number
```

- Visualise the expression of that gene across tumour grades.
  Facet by `sex`.
  Color by `T` score.

```{r}
table_for_ggplot <- exprs_sampleinfo %>% 
  
  
table_for_ggplot
```

```{r}
ggplot(table_for_ggplot, ) +

```

- Draw a heat map of all nine genes.
  Remove labels as well as ticks for the x-axis.
  Remove the panel grid and background.

```{r}
ggplot(exprs_sampleinfo, ) +
  geom_tile() +
    
```

### Bonus point

- In the heat map, group samples by sex.

```{r}
# Copy the code chunk above and tweak the aes(...) contents.

```