---
title: "Example code for introduction to Bioconductor"
author: "Kevin Rue-Albrecht/updated by Carla Cohen"
date: "23/03/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(Biostrings)
library(BSgenome)
library(BSgenome.Hsapiens.UCSC.hg38.masked)
library(rtracklayer)
library(org.Hs.eg.db)
library(biomaRt)
library(SummarizedExperiment)
library(tximeta)
library(tximportData)
library(DESeq2)
library(SingleCellExperiment)
library(ExperimentHub)
library(iSEE)
library(scuttle)
library(scater)
```

# Exercise

## Exploring software packages

- Visit the listing of packages on the Bioconductor [biocViews](https://bioconductor.org/packages/release/BiocViews.html) web page.

- Use the `Autocomplete biocViews search` box in the upper left to identify packages that have been tagged for RNA sequencing analysis.

- Explore other analysis like ChIP-seq, epigenetics, variant annotation, proteomics, single-cell genomics, etc.

- Explore the graph of software packages by expanding and contracting individual terms.

- In the RNA-seq category, find out which of `DESeq2` and `edgeR` is more popular, and go to their landing page.

- Briefly explore the vignette and reference manual links.

When would you consult the vignette? When would the reference manual be helpful?

> Answer:
> 

# Demo

## Bioconductor uses the S4 class system

Define the class.

```{r}
setClass("Person",
  representation(
    name = "character",
    age = "numeric")
  )
```

Define the constructor function.

```{r}
Person <- function(name, age) {
  new("Person", name = name, age = age)
}
```

Use the constructor function to make an object.

```{r}
kevin <- Person(name = "Kevin", age = 21)
kevin
```

Don't access internal slots directly.

```{r}
kevin@age # don't do that
```

The package developers should provide you with accessor functions.

```{r}
get_age <- function(object) {
  object@age
}
```

In which case, you should use them.

```{r}
get_age(kevin) # do that (if possible)
```

# Exercise

## Install Bioconductor

On the teaching cluster, all the packages required for the course are already installed.
As a result, you do not need (and should not) install any package yourself.

> Instead, you should ask the system administrators to install or update packages for you.

To give you an idea, what the system administrators did essentially comes down to:

```{r, include=TRUE, echo=TRUE, eval=FALSE}
!!! DO NOT RUN!!
#install.packages("BiocManager")
#BiocManager::install(c("Biostrings", "S4Vectors"))
```

The package `r BiocStyle::CRANpkg("BiocManager")` is the only official Bioconductor package on CRAN.
It provides functions to install, update, and more generally manage your library of Bioconductor (and CRAN) packages.

The function `BiocManager::install()` can install packages from Bioconductor, CRAN, and even GitHub.
Here, we demonstrate it with a vector of two package names.
However, it can be used to install a single package name as a character value.

- Run the function `BiocManager::valid()`.
  What does it do?

```{r}
BiocManager::valid()
```

> Answer:
> 

# Exercise

## Use Bioconductor packages and help pages

- Load the `Biostrings` package.

```{r}
library(Biostrings)
```

- The package provides the object `DNA_ALPHABET`.
  Print the object in the console and use the help page to explain the contents of the object.

```{r}
DNA_ALPHABET
```
It gives us codes for DNA nucleotides as per: https://www.bioinformatics.org/sms/iupac.html
e.g. S=G or C


```{r, eval=FALSE}
help(DNA_ALPHABET)
```

- The package provides the function `IlluminaQuality()`.
  Run some code from the "Examples" section of the help page, and describe what the function does.
  
```{r, eval=FALSE}
help(IlluminaQuality)
```
  
```{r}
PhredQuality(0:40) #the PhredQuality object is a string of 41 symbols that represent quality scores

```

- Access the package vignette(s).
  How many vignette does the package include?
  
  https://bioconductor.org/packages/release/bioc/html/Biostrings.html

```{r, eval=FALSE}
browseVignettes(package = "Biostrings") #doesn't actually work on the teaching cluster!
#type in R search page instead
```

- Import sequences in the file `TruSeq3-PE-2.fa`.
  What is the class of the object? "DNAStringSet"

?Use the help page to find out how to read in a fasta file.

Reminder: fasta format
>name
AATACAT
>name2
ATCGCGC


```{r}
truseq_adapters <- readDNAStringSet("/project/obds/shared/resources/3_r_stats_genomics/r_bioconductor/TruSeq3-PE-2.fa")
truseq_adapters
class(truseq_adapters)

```
It has created an object of class ""DNAStringSet" which contains the sequence names and sequences. 
# Exercise

## BSgenome packages

- Load the package `BSgenome`.

```{r}
library(BSgenome)
```

- Use the function `BSgenome::available.genomes()`.
  What does it do?

Lists all the available genomes in their package name format.
```{r}
BSgenome::available.genomes()
```

- Load the package `BSgenome.Hsapiens.UCSC.hg38.masked`.
This is the sub-package with the genome you want. 

Masked means that repeats are masked
```{r}
library(BSgenome.Hsapiens.UCSC.hg38.masked)
```

- Assign the genome object provided in the package to a new object called `genome`.

The ones that looks like chr4_GL£&adf_alt etc are partial contigs, we know which chromosome thye are but not exactly where.
If we excluded it, then we would not get reads mapped there. 
Also to do with hypervariable regions in immune genes (the onese that end _alt).There are many versions of HLA-A for example.



```{r}
genome <- BSgenome.Hsapiens.UCSC.hg38.masked
genome
seqnames(genome)

```

An E-box (enhancer box) is a DNA response element found in some eukaryotes that acts as a protein-binding site and has been found to regulate gene expression in neurons, muscles, and other tissues.

The E-box motif is `"CANNTG"`.

- Extract the sequence of chromosome Y from the `genome` object above.

```{r}
chrY <- genome$chrY
```

- Count the number of E-box motifs present on chromosome Y using the appropriate method available in the `Biostrings` package.

```{r}
Ebox <- "CANNTG"
countPattern(Ebox, chrY, fixed = FALSE)
```
No patterns are found containing "N" because htere are not Ns in the genome. 
So we need to include the argument "fixed = FALSE" which will allow it to match N to anything. 

There are lots of other functions
e.g. "translate" to convert DNA to protein sequence.


# Demo

## Creating genomic ranges

```{r}
demo_granges <- GRanges(
  seqnames = c("chr1", "chr2"),
  ranges = IRanges(
    start = c(10, 20),
    end = c(25, 35)),
  metadata1 = c("control", "target"),
  metadata2 = c(1, 2))
demo_granges
```

**What genomics file format does this remind you of?**

> Answer:
> bed

# Demo

## Accessing the contents of GRanges objects

```{r}
seqnames(demo_granges)
```

```{r}
start(demo_granges)
```

```{r}
end(demo_granges)
```

```{r}
seqlevels(demo_granges)
```
These are treated like a factor.

# Demo

## Metadata columns

```{r}
mcols(demo_granges)
```
Each range has some metadata.



```{r}
demo_granges$metadata1
```
Then you can also add metadata to the entire object. (Currently an empty list)

```{r}
metadata(demo_granges)
```

# Demo

## DataFrame

```{r}
library(S4Vectors)
demo_DataFrame <- DataFrame(
    column1 = c("A", "B"),
    column2 = c(1, 2)
)
demo_DataFrame
```
This is a different style of df where there is a data table with an associated metadata table.


```{r}
#add the metadata
mcols(demo_DataFrame) <- DataFrame(
  column_type = c("character", "numeric")
)
mcols(demo_DataFrame)
```

# Exercise

## rtracklayer
This package is used to import files such as gtf and bed that can be used by GenomicRanges.

- Load the `rtracklayer` package.

```{r}
library(rtracklayer)
```

- Import the file `Homo_sapiens.GRCh38.104.gtf.gz` using the function `rtracklayer::import()`.
  Assign value to an object called `gtf_data`.

```{r}
gtf_data <- rtracklayer::import("/project/obds/shared/resources/3_r_stats_genomics/r_bioconductor/Homo_sapiens.GRCh38.104.gtf.gz")
gtf_data

```

- What are the metadata columns available in the object?

```{r}
mcols(gtf_data) #gives you the whole DataFrame
colnames(mcols(gtf_data)) #shows just the column names
```

- Use the function `subset()` to filter the annotations for the gene `ACTB`.

```{r}
head(gtf_ACTB$gene_id)

gtf_ACTB <- subset(gtf_data, gene_name == "ACTB")
gtf_ACTB
```

- How many transcripts do the annotations describe for that gene?

```{r}
#how many transcript identifiers are there?
gtf_ACTB$transcript_id %>%  #extract the transcript IDs
  unique() %>% #select the unique values
  setdiff(NA) %>% #remove the NAs, can also use na.omit
  length() #count them

#in the gtf file each transcript has multiple entries for gene, each exon, coding sequences etc.



```

- Subset the annotations to exons, and use the `split()` function to separate exons from different transcripts.
  What is the class of the object that is returned?
  GRangesList object.

```{r}
gtf_ACTB$type
gtf_ACTB_exons <- subset(gtf_ACTB, type == "exon")  #select the exons of ACTB
gtf_ACTB_exons$transcript_id #look for the transcript_ids
gtf_ACTB_transcripts <- split(gtf_ACTB_exons, gtf_ACTB_exons$transcript_id) #split the exons by transcript id
gtf_ACTB_transcripts #have a look at the output
str(gtf_ACTB_transcripts) #view the structure
gtf_ACTB_transcripts$ENST00000414620 #view the ranges for one transcript_id
```

# Exercise

## Annotation packages

Each annotation packages contains a database of information.

The `org.db` family of packages provide annotations that link  several types of identifiers for genes, proteins, and pathways (e.g. KEGG, Gene Ontology).

- Load the package `org.Hs.eg.db`.

```{r}
library(org.Hs.eg.db)
```

Each `OrgDb` package contains an object of the same name as the package.

- What is the class of the object?
OrgDb object

```{r}
org.Hs.eg.db
```

- Use the `columns()` method to discover which sorts of annotations can be extracted
from the object.

```{r, include=TRUE}
columns(org.Hs.eg.db)
```

- Use the `keytypes()` method to discover which columns can be used to query information.

```{r}
keytypes(org.Hs.eg.db)
```
eg ENSEMBL contains EnsemblID
GO gene ontoloty
Symbol - gene symbol 
PMID pub med identifier of papers mentioning this gene

In this case columns and keytypes are the same list (but not always).

- Use the `keys()` method to get the first six gene symbols in the database.
  Store as `human_symbols`.

```{r}
human_symbols <- head(keys(org.Hs.eg.db, keytype = "SYMBOL"), n=6)
human_symbols
```

- Use the `AnnotationDbi::select()` method to fetch the Entrez and Ensembl identifiers for those gene symbols.

```{r}
AnnotationDbi::select(org.Hs.eg.db, #search this db
                      keys =human_symbols, #this is what we are looking for
                      columns = c("ENTREZID", "ENSEMBL"), #this is what we want out
                      keytype = "SYMBOL") #this is the type of query we are giving

```

  + What happens if you do not specify `AnnotationDbi`?

> Answer:
> It might use select from another package

- Use the `mapIds()` method to get the Ensembl identifier for those gene symbols.

This is similar to above but the output is a named vector rather than a table.
So useful if you want to replace ensemble ID with symbol.


```{r}
mapIds(org.Hs.eg.db, keys = human_symbols, column = "ENSEMBL", keytype = "SYMBOL")
```

# Exercise

## biomaRt

- Load the package `biomaRt`.

```{r}
library(biomaRt)
```

- Use the `listMarts()` function to list the BioMart databases to which the `biomaRt` package can connect.
This contacts the online db and prints the answer here.
We will use Ensemble Genes today. ENSEMBL_MART_ENSEMBL

```{r}
listMarts()
```

- Use the `listEnsemblArchives()` function to lists the available archived versions of Ensembl.
The Ensembl website also has versions that are released periodically.

```{r}
listEnsemblArchives()
```

- Use the `useMart()` function to create an object named `mart` using the `"ENSEMBL_MART_ENSEMBL"` BioMart database and set the host option to `"https://may2021.archive.ensembl.org"`.
  Why would one do that in practice?

```{r}
mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", host = "https://may2021.archive.ensembl.org") 
#leave out dataset for now as we don't know which one we want

mart
```

> Answer:
> 

- Use the `listDatasets()` function to list the datasets available in the selected BioMart database.
  What is the identifier of the database that contains information for _Homo sapiens_?

```{r}
listDatasets(mart)
View(listDatasets(mart)) #view the table and easily search for human
#hsapiens_gene_ensembl

```

- Use again the `useMart()` function, and replace the `mart` object by a new one that points to information for _Homo sapiens_.

```{r}
mart <- useMart(biomart = "ENSEMBL_MART_ENSEMBL", 
                dataset = "hsapiens_gene_ensembl",
                host = "https://may2021.archive.ensembl.org")
mart
```

- Use the `listAttributes()` function to list the fields of information available in the dataset.
Attributes are the equivalent to "columns" in the example above.

```{r}
listAttributes(mart)
View(listAttributes(mart))
```

- Use the `getBM()` function to fetch the chromosome, start and end positions, and strand information for the gene symbols that you stored earlier as `human_symbols`.

```{r}
getBM(attributes = c("hgnc_symbol", "chromosome_name", "start_position", "end_position", "strand"), #what output do you want?
      filters = "hgnc_symbol", #where are your search terms coming from
      values = human_symbols, #what are your search terms?
      mart = mart) #what is the database you are searching?
```

# Demo

## EnsDb packages

```{r}
library(EnsDb.Hsapiens.v86)
```

```{r}
class(EnsDb.Hsapiens.v86)
```

```{r}
columns(EnsDb.Hsapiens.v86)
```

# Demo

## TxDb packages

```{r}
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
```

```{r}
class(TxDb.Hsapiens.UCSC.hg38.knownGene)
```

```{r}
columns(TxDb.Hsapiens.UCSC.hg38.knownGene)
```

```{r}
keytypes(TxDb.Hsapiens.UCSC.hg38.knownGene)
```

```{r}
keys(TxDb.Hsapiens.UCSC.hg38.knownGene, "GENEID") %>% head()
```

# Exercise

## Creating a SummarizedExperiment object
demo_count_matrix = Making up some data to represent the experiment
demo_rowdata is the gene metadata
coldata is the sample metadata eg control, treated

demo_se is created from these three matrices using the command SummarizedExperiment. 

Usually you would read these files into R from your our experimental output/spreadsheets etc. 

```{r, include=TRUE, results='hide'}
library(SummarizedExperiment)
demo_count_matrix <- matrix(
  data = c(1, 2, 3, 4, 5, 6, 7, 8), nrow = 4, ncol = 2,
  dimnames = list(
    paste0("gene", 1:4),
    paste0("sample", 1:2)
  ))
demo_rowdata <- DataFrame(
  symbol = head(letters, 4),
  length = c(1234, 5678)
  )
demo_coldata <- DataFrame(
  condition = c("control", "treated"),
  age = c(18, 20)
  )
demo_se <- SummarizedExperiment(
  assays = list(
    counts = demo_count_matrix),
  rowData = demo_rowdata,
  colData = demo_coldata)
```

```{r, include=TRUE}
demo_se #summary view of object
str(demo_se) #more verbose output
```

Use the methods `assays()`, `assayNames()` `assay()`, `colData()`, `rowData()`, `mcols()` and `metadata()` on the object `demo_se`.

```{r}
assays(demo_se) #extracts all the assays as a list
assayNames(demo_se) #what are the names of the assays in the object (here there is just one called counts)
assay(demo_se, "counts") #view the assay data for one assay
#there will only ever be one colData and rowData as this doeos not change as the data is analysed (e.g. age of patient)
colData(demo_se)
rowData(demo_se)
#can also add additional information to this metadata
colData(demo_se)$score <- c(100, 200)
mcols(demo_se)
#you can also add metadata for the whole experiment (e.g. the name of the experiment) rather than the samples therein. 
metadata(demo_se) <- list(experiment_name = "Demonstration")
demo_se

```

What do those functions do?

> Answer:
> 

# Exercise

## Assembling a SummarizedExperiment object

- Import the matrix in the file `counts.csv`.
  Call it `assay_counts`.

```{r}
assay_counts <- read.table("/project/obds/shared/resources/3_r_stats_genomics/r_bioconductor/summarized_experiment/counts.csv",
                           row.names = 1, sep = ",", header = TRUE)
head(assay_counts)
#colnames(assay_counts) <- assay_counts[1,]

#this is a df, convert to matrix
assay_counts <- as.matrix(assay_counts)

class(assay_counts)
head(assay_counts)
```

- Import the data frames in the files `sample_metadata.csv` and `gene_metadata.csv`.
  Call them `sample_metadata` and `gene_metadata`.

```{r}
sample_metadata <- read.csv("/project/obds/shared/resources/3_r_stats_genomics/r_bioconductor/summarized_experiment/sample_metadata.csv")
head(sample_metadata)
```

```{r}
gene_metadata <- read.csv("/project/obds/shared/resources/3_r_stats_genomics/r_bioconductor/summarized_experiment/gene_metadata.csv")
head(gene_metadata)
```

- Use the function `SummarizedExperiment()` to combine those three objects into a single `SummarizedExperiment` object.
  Call it `demo_se`.
  
  + Assign the matrix to an assay named `counts`.

This is a really important step because if you set up your experiment wrong then you might compare the wrong things in the analysis!

```{r}
demo_se <- SummarizedExperiment(assays = list(counts = assay_counts),
  rowData = gene_metadata,
  colData = sample_metadata)
  
demo_se

assays(demo_se)
```

# Demo

## Inheritance and coercion

```{r}
demo_dataframe <- data.frame(A = c(1, 2), B = c(3, 4))
demo_dataframe
```

```{r}
demo_DataFrame <- as(demo_dataframe, "DataFrame")
demo_DataFrame
```

# Exercise

## SummarizedExperiment extensions

- Load the package `DESeq2`.

```{r}
library(DESeq2)
```

- Convert your earlier `demo_se` to a `DESeqDataSet` object, using the `as()` function.
  Call the object `demo_deseq`.

```{r}
demo_deseq <- as(demo_se, "DESeqDataSet")
demo_deseq
```

- What are the slots of information present in the `DESeqDataSet` object?
  Use the function `slotNames()`.

```{r}
slotNames(demo_deseq)
```


- Which slots are new compared to the `SummarizedExperiment` object?

```{r}
slotNames(demo_se)
```
We now have the options design, dispersionFunction, rowRanges.


- Which components of the `DESeqDataSet` object can you access using accessor functions (e.g., `assays()`) ?
  Do all of those accessors work on the original `demo_se` object?

```{r}

design(demo_deseq) #currently null as not yet defined
assays(demo_deseq) #same as for the se object

```

> Answer:
> 

# Exercise

## SingleCellExperiment

The class name `SingleCellExperiment` is only a name.
While developed with single-cell genomics in mind, it is not limited to that use.
Remember that classes are simply defined by their slots and methods.
If the slots and methods are suitable to store and process another type of data, then go for it!

- Install and load the package `SingleCellExperiment`.

```{r}
library(SingleCellExperiment)
```

- Convert your earlier `demo_se` to a `SingleCellExperiment` object, using the `as()` function.
  Call the object `demo_sce`.

```{r}
demo_sce <- as(demo_se, "SingleCellExperiment")
```

- Display the object.
  What additional information can the `SingleCellExperiment` class store, relative to the `SummarizedExperiment` class?

```{r}
demo_sce
```

> Answer:
> 

- Use the methods `reducedDimNames()`, `reducedDims()`, and `reducedDim()` on the object `demo_sce`.
  What do they do?

```{r}
reducedDimNames(demo_sce)
reducedDims(demo_sce)
reducedDim(demo_sce)
```

> Answer:
> 

- **Bonus point:** Use the method `reducedDim()` to store a new dimensionality reduction matrix called `"PCA"` in the object.
  Display and inspect the updated object.

```{r}
pca_object <- prcomp(t(assay(demo_sce)))
reducedDim(demo_sce, "PCA") <- pca_object$x
demo_sce
```

# Demo

## The ExperimentHub package and resources

```{r, include=TRUE}
library(ExperimentHub)
ehub <- ExperimentHub()
ehub
```

```{r}
query(ehub, c("SingleCellExperiment"))
```

```{r, include=TRUE}
eh1433 <- ehub[["EH1433"]]
eh1433
```

# Demo

## Description of a dataset in the ExperimentHub

```{r}
ehub["EH1433"]
```

# Exercise

## Interactive visualisation using iSEE

- Load the package `iSEE`.

```{r}
library(iSEE)
```

- Use the `iSEE()` function on the `SingleCellExperiment` object `eh1433`.

```{r, include=TRUE, eval=FALSE}
iSEE(eh1433)
```

- Subset the object `eh1433` to remove cells where the sum of counts is `NA`.

```{r}
eh1433 <- eh1433[, !is.na(colSums(assay(eh1433)))]
eh1433
```

- Use the packages `scuttle` and `scater` to compute and add a PCA result to the object `eh1433`

```{r}
library(scuttle)
library(scater)
eh1433 <- logNormCounts(eh1433)
eh1433 <- runPCA(eh1433)
```

- Launch the `iSEE` GUI again.
  What changed?

```{r, eval=FALSE}
iSEE(eh1433)
```

# Advanced

## BiocManager repositories

- Run `options("repos")`.
  What do you see?

```{r}
options("repos")
```

The function `options()` can be used to get or set global options of the current <i class="fab fa-r-project"></i> session.

- Run `BiocManager::repositories()`.

```{r}
BiocManager::repositories()
```

The function reports the URLs from which to install Bioconductor and CRAN packages.

- Set the `repos` option to the value of `BiocManager::repositories()`.

```{r}
options(repos = BiocManager::repositories())
```

- Run `options("repos")` again.
  What do you see now?
  What does it mean?

```{r}
options("repos")
```

> Answer:
> 

# Advanced

## Import SummarizedExperiment objects using tximeta

- Load the packages `tximeta` and `tximportData`.

```{r}
library(tximeta)
```

The help page of the `tximeta()` function refers to an example output of the Salmon program that is installed on your computer at `system.file("extdata/salmon_dm/SRR1197474/quant.sf", package="tximportData")`.

```{r, eval=FALSE}
help(tximeta)
```

- Navigate to that location in the file explorer of your respective operating system, and examine its contents, in a text editor or your terminal application.

```{r, eval=FALSE}
file.edit(system.file("extdata/salmon_dm/SRR1197474/quant.sf", package="tximportData"))
```

- Run the rest of the example code, and discuss the use of a linked transcriptome with `tximeta` (read the help page of `makeLinkedTxome()` and set `write=TRUE` instead of using the example code as-is).

```{r}
dir <- system.file("extdata/salmon_dm", package="tximportData")
files <- file.path(dir, "SRR1197474", "quant.sf") 
coldata <- data.frame(files, names="SRR1197474", condition="A", stringsAsFactors=FALSE)
```

```{r}
se <- tximeta(coldata)
se
```

```{r}
indexDir <- file.path(dir, "Dm.BDGP6.22.98_salmon-0.14.1")
fastaFTP <- c("ftp://ftp.ensembl.org/pub/release-98/fasta/drosophila_melanogaster/cdna/Drosophila_melanogaster.BDGP6.22.cdna.all.fa.gz",
              "ftp://ftp.ensembl.org/pub/release-98/fasta/drosophila_melanogaster/ncrna/Drosophila_melanogaster.BDGP6.22.ncrna.fa.gz")
gtfPath <- file.path(dir, "Drosophila_melanogaster.BDGP6.22.98.gtf.gz")
makeLinkedTxome(indexDir=indexDir, source="Ensembl", organism="Drosophila melanogaster",
                release="98", genome="BDGP6.22", fasta=fastaFTP, gtf=gtfPath, write=TRUE, jsonFile = "tximeta.json")
se <- tximeta(coldata)
se
```

- What other types of quantification programs does `tximeta` support?
  Where do you find that information?

```{r, eval=FALSE}
help(tximport)
```

> Answer: